---
title: "Number of L2G-prioritized genes"
output: html_document
---


```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(annotables)
options(stringsAsFactors = F)
theme_set(theme_bw())

library(tidyverse)
theme_set(theme_bw())

otgdir = "/Users/jeremys/work/otgenetics"
df = read_tsv("/Users/jeremys/work/otgenetics/otg/l2g.full.tsv.gz", col_types = cols_only(study_id = "c", gene_id = "c", y_proba = "d")) %>%
  rename(l2g = y_proba)
studies = read_tsv("/Users/jeremys/work/otgenetics/otg/studies.tsv.gz")

df = df %>%
  left_join(studies %>% select(study_id, trait_reported, trait_category, has_sumstats)) %>%
  left_join(grch38 %>% select(gene_id = ensgene, symbol) %>% filter(!duplicated(gene_id)))

trait_category_counts = studies %>%
  group_by(trait_category) %>%
  summarise(n = n())
```

Total number of unique genes with high L2G score in any study.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
num = df %>% filter(l2g > 0.5) %>% .$gene_id %>% unique() %>% length()
print(sprintf("L2G > 0.5: %d", num))

num = df %>% filter(has_sumstats, l2g > 0.5) %>% .$gene_id %>% unique() %>% length()
print(sprintf("L2G > 0.5 (has sumstats): %d", num))

num = df %>% filter(l2g > 0.8) %>% .$gene_id %>% unique() %>% length()
print(sprintf("L2G > 0.8: %d", num))

num = df %>% filter(has_sumstats, l2g > 0.8) %>% .$gene_id %>% unique() %>% length()
print(sprintf("L2G > 0.8 (has sumstats): %d", num))

```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
# Save a list of all genes with the max L2G score per gene
max_l2g = df %>%
  arrange(desc(l2g)) %>%
  filter(!duplicated(gene_id)) %>%
  select(study_id, gene_id, symbol, l2g, trait_reported, trait_category)
write_tsv(max_l2g, path = file.path(otgdir, "misc/l2g_summary/max_l2g_per_gene.tsv"), na = "")
```

## Number of studies in which a given gene has high L2G score
Notice log-scaled axis.

```{r, warning=FALSE, message=FALSE, echo=FALSE}
df %>%
  filter(l2g > 0.5) %>%
  group_by(gene_id) %>%
  filter(!duplicated(study_id)) %>%
  summarise(num_studies = n()) %>%
  ggplot(aes(x=num_studies)) +
  geom_histogram(bins=50) +
  scale_y_log10() +
  ggtitle("L2G > 0.5") +
  xlab("Number of genes prioritised in study") + ylab("Number of studies")
```

```{r, warning=FALSE, message=FALSE, echo=FALSE}
df %>%
  filter(l2g > 0.8) %>%
  group_by(gene_id) %>%
  filter(!duplicated(study_id)) %>%
  summarise(num_studies = n()) %>%
  ggplot(aes(x=num_studies)) +
  geom_histogram(bins=50) +
  scale_y_log10() +
  ggtitle("L2G > 0.8") +
  xlab("Number of genes prioritised in study") + ylab("Number of studies")
```

## Genes with L2G > 0.5 in the most studies

```{r, warning=FALSE, message=FALSE, echo=FALSE}
df %>%
  filter(l2g > 0.5) %>%
  group_by(gene_id) %>%
  filter(!duplicated(study_id)) %>%
  summarise(num_studies = n(),
            symbol = first(symbol)) %>%
  arrange(desc(num_studies)) %>%
  select(symbol, num_studies)
```

AC002094.3 overlaps gene VTN, and seems to be a trans-pQTL for a large number of proteins in the Emilsson study.

Without the Emilsson study:

```{r, warning=FALSE, message=FALSE, echo=FALSE}
study_counts = df %>%
  filter(l2g > 0.5) %>%
  filter(!grepl("GCST006585_", study_id)) %>%
  group_by(gene_id) %>%
  filter(!duplicated(study_id)) %>%
  summarise(symbol = first(symbol),
            num_studies = n())

study_counts %>%
  arrange(desc(num_studies)) %>%
  write_tsv(path = file.path(otgdir, "misc/l2g_summary/num_studies_per_gene_l2g_gt_0.5.tsv"), na = "")

study_counts %>%
  arrange(desc(num_studies)) %>%
  select(symbol, num_studies)
```

## Studies prioritising the most genes
(excluding Emilsson)

```{r, warning=FALSE, message=FALSE, echo=FALSE}
gt_0.5 = df %>%
  filter(l2g > 0.5) %>%
  filter(!grepl("GCST006585_", study_id)) %>%
  filter(!duplicated(gene_id)) %>%
  group_by(study_id) %>%
  summarise(num_genes = n()) %>%
  arrange(desc(num_genes))

gt_0.8 = df %>%
  filter(l2g > 0.8) %>%
  filter(!grepl("GCST006585_", study_id)) %>%
  filter(!duplicated(gene_id)) %>%
  group_by(study_id) %>%
  summarise(num_genes = n()) %>%
  arrange(desc(num_genes))

gene_counts = studies %>%
  left_join(gt_0.5 %>% rename('L2G>0.5' = num_genes), by="study_id") %>%
  left_join(gt_0.8 %>% rename('L2G>0.8' = num_genes), by="study_id")
gene_counts$`L2G>0.5`[is.na(gene_counts$`L2G>0.5`)] = 0
gene_counts$`L2G>0.8`[is.na(gene_counts$`L2G>0.8`)] = 0

gene_counts %>%
  arrange(desc(`L2G>0.5`)) %>%
  select(study_id, trait_reported, `L2G>0.5`, `L2G>0.8`, pub_title) %>%
  head(10)
```

Studies prioritising the most genes, also excluding anthropometric measurements:

```{r, warning=FALSE, message=FALSE, echo=FALSE}
gene_counts %>%
  arrange(desc(`L2G>0.5`)) %>%
  filter(!grepl("Anthropometric", trait_category, ignore.case = T)) %>%
  filter(!grepl("Measurement", trait_category, ignore.case = T)) %>%
  filter(!grepl("UKB", pub_author, ignore.case = T)) %>%
  select(study_id, trait_reported, `L2G>0.5`, `L2G>0.8`, pub_title) %>%
  head(10)

```


```{r, warning=FALSE, message=FALSE, echo=FALSE}

# Number of unique genes with L2G score per study
# df %>%
#   group_by(study_id) %>%
#   filter(!duplicated(gene_id)) %>%
#   summarise(num_genes = n()) %>%
#   ggplot(aes(x=num_genes)) +
#   geom_histogram() +
#   scale_y_log10()

# Number of unique genes with L2G > 0.5 per study
num_unique_genes = df %>%
  group_by(study_id) %>%
  filter(l2g > 0.5, !duplicated(gene_id)) %>%
  summarise(num_genes_prioritised = n())

num_unique_genes %>%
  ggplot(aes(x=num_genes_prioritised)) +
  geom_histogram() +
  scale_y_log10() + scale_x_log10() +
  ggtitle("Number of studies prioritising (L2G > 0.5) a given number of genes") +
  ylab("Number of studies") + xlab("Number of genes prioritised")

gene_counts %>%
  arrange(desc(`L2G>0.5`)) %>%
  write_tsv(path = file.path(otgdir, "misc/l2g_summary/num_genes_per_study_l2g_gt_0.5.tsv"), na = "")
```

